New lower bounds for Schur and weak Schur numbers

### 2021-04-15 article/article.tex
reorganisation sat / search space
Bibliographie par ordre d'apparition
introduction
restructuration + indicatrice plus jolie

\usepackage{amsmath}
\usepackage{amssymb}

@@

This article provides new lower bounds for both Schur and weak Schur numbers. These results were obtained by continuing on
Rowley's "templates"-based approach for Ramsey and Schur numbers. Finding suitable templates allows us to apply Rowley's construction
to get explicit partitions improving lower bounds as well as the growth rate of both Schur numbers and Ramsey numbers \(R_n(3)\).
We also developed a new method to improve lower bounds on weak Schur numbers. Furthermore, this paper tries to analyze former
works on the subject based on the principle that good partitions into \(n+1\) subsets start with good partitions
into \(n\) subsets. We show that exceeding the previous lower bound \(\WS (6) \geqslant 582\) is impossible with such an assumption upon
imposing certain conditions on the good 5-subsets partition. The new lower bounds include \(S(9) \geqslant 17803\), \(S(10) \geqslant 60948\),
\(\WS (9) \geqslant 22536 \) and \(\WS (10) \geqslant 71214 \).

@@

\include{state_of_the_art.tex}

\include{definitions_notations.tex}

@@

\include{sat.tex}

@@

\par
These new results come from an extension of Rowley's template-based approach for Ramsey graphs and Schur numbers which is relatively new.
Therefore, we would not be surprised if lower bounds are later improved using better templates. Moreover, studying specifically \(S^+(n)\) and
\(\WS^+(n)\) might be of interest as they are closely related to Schur and weak Schur numbers.

@@

\par
We would first like to thank our professors, Joanna Tomasik and Arpad Rimmel, who guided and advised
us during the whole research and writing. Last but not least, we would like to address warm thanks
to Fred Rowley who kindly answered our many questions and with who it was a pleasure to exchange.

\bibliography{biblio}
\bibliographystyle{plain}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{dsfont}

@@

This article provides new lower bounds for both Schur and weak Schur numbers. These results were obtained by continuing on
Rowley's "template"-based approach for Schur and Ramsey numbers. Finding suitable templates allows us to apply Rowley's construction
to get explicit partitions improving lower bounds as well as the growth rate for both Schur numbers and Ramsey numbers \(R_n(3)\).
We also developed a new method to improve lower bounds on weak Schur numbers. Furthermore, this paper tries to analyze former
works on the subject based on the principle that good partitions into \(n+1\) subsets start with good partitions
into \(n\) subsets. We show that exceeding the previous lower bound \(\WS (6) \geqslant 582\) is impossible with such an assumption upon
imposing certain conditions on the good 5-subsets partition. The new lower bounds include \(S(9) \geqslant 17803\), \(S(10) \geqslant 60948\),
\(\WS (9) \geqslant 22536 \) and \(\WS (10) \geqslant 71214 \).

@@

% \include{definitions_notations.tex}

@@

\include{search_space.tex}

@@

\qquad These new results come from an extension of Rowley's template-based approach for Ramsey graphs and Schur numbers which is relatively new.
Therefore, we would not be surprised if lower bounds are later improved using better templates. Moreover, studying specifically \(S^+(n)\) and
\(\WS^+(n)\) might be of interest as they are closely related to Schur and weak Schur numbers.

@@

\qquad We would first like to thank our professors, Joanna Tomasik and Arpad Rimmel, who guided and advised
us during the whole research and writing. Last but not least, we would like to address warm thanks
to Fred Rowley who kindly answered our many questions and with who it was a pleasure to exchange.

\bibliography{biblio}
\bibliographystyle{ieeetr}

### 2021-04-15 article/introduction.tex
restructuration + indicatrice plus jolie

We are interested in partioning the set of integers \(\{1, ..., p\}\) in \(n\) subsets such that there is no subset 
containing three integers \(x\), \(y\) and \(z\) verifying \(x + y = z\). We say the subsets are sum-free. If we 
add the hypothesis \(x \neq y\), we say the subsets are weakly sum-free. The greatest \(p\) for which a 
partition into \(n\) sum-free exists is called the \(n^{\text{th}}\) Schur number and is denoted \(S(n)\). 
\cite{Schur1917} Likewise for weakly sum-free partitions we define \(WS(n)\) the \(n^{\text{th}}\) weak Schur 
number. \cite{Irving1973} Only the first values of these sequences are known.

\par Before Rowley's "template"-based approach for Schur and Ramsey numbers \cite{RowleyRamsey}, the 
previous generic construction for Schur numbers was given by Abott and Hanson \cite{AbottHanson} in 1972 
with a recursive construction. It gave the best lower bounds for all sufficiently large bounds. No equivalent 
was known for weak Schur numbers and as a result the best known partitions for large weak Schur numbers 
were actually sum-free partitions. As for smaller numbers, the best lower bounds were obtained by conducting 
a computer search. This search for weakly sum-free partitions relied on the recursive assumption that a good 
weakly sum-free partition into \(n+1\) colors starts with a good weakly sum-freepartition into \(n\) colors. In 
2020, Rowley introduced the notion of template for Schur and Ramsey numbers which generalizes Abott and 
Hanson's construction and gives new lower bounds (and inequalties) for Schur numbers. Rowley also gives two 
inequalities for weak Schur numbers that yield significant improvements over previous lower bounds which 
besides are higher than the lower bounds on Schur numbers. \cite{RowleyWS}

\renewcommand{\arraystretch}{1.5}
\setlength{\tabcolsep}{2pt}

\begin{center}
\begin{tabular}{|*{13}{c|}}
    \hline
    \(k\) & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 \\
    \hline
    \(S(n)\) & & & & & & & & & & & & \\
    old & 1 & 4 & 13 & 44 & 160 \cite{Heule2017} & 536 \cite{Fredricksen} & 1680 \cite{Fredricksen} & 5041 
    \cite{ELIAHOU2012175} & 15124 \cite{ELIAHOU2012175} & 51120 \cite{AbbottHanson} & 172216 
    \cite{AbbottHanson} & 575664 \cite{AbbottHanson} \\
    Rowley & & & & & & & & 5286 & 17694 & 60320 & 201696 & 631840 \\
    new & & & & & & & & & 17803 & 60948 & 203828 & 638548 \\
    \hline
    \(\WS(n)\) & & & & & & & & & & & & \\
    old & 2 & 8 & 23 & 66 & 196 \cite{ELIAHOU2012175} & 582 \cite{EliahouBook} & 1740 
    \cite{Rafilipojaona} & 5201 \cite{Rafilipojaona} & 15596 \cite{Rafilipojaona} & 51520 & 172216 & 575664 \\
    Rowley & & & & & & 642 & 2146 & 6976 & 21848 & 70778 & 241282 & 806786 \\
    new & & & & & & & & & 22536 & 71214 & 243794 & 815314 \\
    \hline
\end{tabular}
\end{center}

\par The main contribution of this article is a generalization of the concept of template to weak Schur numbers. 
This gives new lower bounds (and inequalities) for weak Schur numbers. This construction also includes as a 
special case an analogous for weak Schur numbers of Abott and Hanson's construction.

\par We first explain Rowley's template-based construction in the context of Schur numbers and then give 
new templates, thus providing new lowers bounds and inequalities as well as showing that the growth rates 
of both Schur and Ramsey numbers exceed 3.28. Then, we generalize the concept of templates to weak Schur 
numbers and provide new lower bounds for weak Schur numbers. Finally, we analyze the significant difference 
between new lower bounds obtain with templates and the former lower bounds obtained by computer search 
and we provide evidence that the main assumption made in those articles removes the optimal partitions from 
the search.

>>>>

We are interested in partioning the set of integers \(\{1, ..., p\}\) in \(n\) subsets such that there is no subset 
containing three integers \(x\), \(y\) and \(z\) verifying \(x + y = z\). We say these subsets are sum-free. If we 
add the hypothesis \(x \neq y\), we say the subsets are weakly sum-free. The greatest \(p\) for which a 
partition into \(n\) sum-free exists is called the \(n^{\text{th}}\) Schur number and is denoted \(S(n)\) 
\cite{Schur1917}. Likewise for weakly sum-free partitions we define \(WS(n)\) the \(n^{\text{th}}\) weak Schur 
number \cite{Irving1973}. Only the first values of these sequences are known. This article concerns the lower 
bounds of these numbers.

\par Before Rowley's "template"-based approach for Schur and Ramsey numbers \cite{RowleyRamsey}, the 
previous generic construction for Schur numbers was given by Abbott and Hanson \cite{AbbottHanson} in 1972 
with a recursive construction. It gave the best lower bounds for all sufficiently large bounds. No equivalent 
was known for weak Schur numbers and as a result the best known partitions for large weak Schur numbers 
did not utilize the weakly sum-free hypothesis. As for smaller numbers, the best lower bounds were obtained 
by conducting a computer search. This search for weakly sum-free partitions relied on the recursive assumption 
that a good weakly sum-free partition into \(n+1\) colors starts with a good weakly sum-free partition into \(n\) 
colors. In 2020, Rowley introduced the notion of template for Schur and Ramsey numbers which generalizes 
Abbott and Hanson's construction and gives new lower bounds (and inequalties) for Schur numbers. Rowley also 
gives two inequalities for weak Schur numbers that yield significant improvements over previous lower bounds 
which besides do utilize the weakly sum-free hypothesis \cite{RowleyWS}.

\renewcommand{\arraystretch}{1.5}

\begin{center}

\textbf{Table 1 - Comparison of lower bounds for Schur numbers}

\begin{tabular}{|*{13}{c|}}
    \hline
    \(n\) & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 \\
    \hline
    old & 1 & 4 & 13 & 44 & 160 & 536 & 1680 & 5041 & 15124 & 51120 & 172216 & 575664 \\
      & & & & & \cite{Heule2017} & \cite{Fredricksen} & \cite{Fredricksen} & \cite{ELIAHOU2012175} & 
    \cite{ELIAHOU2012175} & \cite{AbbottHanson} & \cite{AbbottHanson} & \cite{AbbottHanson} \\
    \hline
    Rowley & & & & & & & & 5286 & 17694 & 60320 & 201696 & 631840 \\
    \hline
    new & & & & & & & & & 17803 & 60948 & 203828 & 638548 \\
    \hline
\end{tabular}

\vspace{3ex}
\textbf{Table 2 - Comparison of lower bounds for weak Schur numbers}

\begin{tabular}{|*{13}{c|}}
    \hline
    \(n\) & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 \\
    \hline
    old & 2 & 8 & 23 & 66 & 196 & 582 & 1740 & 5201 & 15596 & 51520 & 172216 & 575664 \\
      & & & & & \cite{ELIAHOU2012175} &\cite{EliahouBook} & \cite{Rafilipojaona} & \cite{Rafilipojaona} & 
    \cite{Rafilipojaona} & \cite{AbbottHanson} & \cite{AbbottHanson} & \cite{AbbottHanson} \\
    \hline
    Rowley & & & & & & 642 & 2146 & 6976 & 21848 & 70778 & 241282 & 806786 \\
    \hline
    new & & & & & & & & & 22536 & 71214 & 243794 & 815314 \\
    \hline
\end{tabular}

\end{center}

\par The main contribution of this article is a generalization of the concept of template to weak Schur numbers. 
This gives new lower bounds (and inequalities) for weak Schur numbers. This construction also includes as a 
special case an analogous for weak Schur numbers of Abbott and Hanson's construction.

\par We first explain Rowley's template-based construction in the context of Schur numbers and then give 
new templates, thus providing new lowers bounds and inequalities as well as showing that the growth rates 
for both Schur and Ramsey numbers exceed 3.28. Then, we generalize the concept of templates to weak Schur 
numbers and provide new lower bounds for weak Schur numbers. Finally, we analyze the significant difference 
between new lower bounds obtain with templates and the former lower bounds obtained by computer search 
and we provide evidence which indicate that the main assumption made in those articles removes the optimal 
partitions from the search space.

\begin{notation}
We denote by \([\![1,p]\!]\) the set of integers \(\{1, 2, ..., p\}\).
\end{notation}

\begin{notation}
For a partition of \([\![1, p]\!]\) in \(n\) subsets, we generally denote these subsets \(A_1, ..., A_n\). We 
also denote \(m_i = \min(A_i)\).
By ordering the subsets, we mean assuming that \(m_1 < ... < m_n\). However, if not specified we do not 
make this hypothesis since we do not always consider partitions in which every subset plays a symmetric role.
\end{notation}

\begin{definition}
We sometimes refer to a partitition as a colouring. The colouring associated to a partition \(A_1, ..., A_n\) of
\([\![1, p]\!]\) is the function \(f\) such that \(\forall x \in [\![1, p]\!], x \in A_{f(x)}\). Likewise, the partition 
associated to a colouring \(f\) of \([\![1, p]\!]\) with \(n\) colors is \(\forall c \in [\![1, n]\!], A_c = f^{-1}(c)\).
\end{definition}

### 2021-04-15 article/schur.tex
restructuration + indicatrice plus jolie

\qquad In this section, we use Rowley's constructions \cite{RowleyRamsey} in the context of Schur numbers. To improve lower bounds
for Ramsay's numbers, Rowley introduces some sum-free partitions verifying some additional properties which can be extended using a 
method which generalizes Abbott and Hanson's  construction \cite{AbbottHanson}. Rowley named these partitions "templates", and 
we will keep this name in the entire article. We find suitable templates and use them to find new lower bounds for Schur numbers.

@@

\subsection{Inequalities using \(S^+\)}

The main result on \(S^+\) follows. It allows us to improve lower bounds on Schur numbers by computing \(S^+\).

@@

>>>>

\qquad In this section, we use Rowley's constructions \cite{RowleyRamsey} in the context of Schur numbers. To improve lower bounds
for Ramsay's numbers, Rowley introduces some sum-free partitions verifying some additional properties which can be extended using a 
method which generalizes Abbott and Hanson's  construction \cite{AbbottHanson}. Rowley named these partitions "templates", and 
we will keep this name in the entire article. We find suitable templates and use them to provide new lower bounds for Schur numbers.

@@

\subsection{Construction of Schur partitions using SF-templates}

\qquad The main result on SF-templates follows. It allows us to improve lower bounds on Schur numbers by finding 
suitable SF-templates.

@@

\subsection{Conclusion on SF-templates}

### 2021-04-15 article/search_space.tex
restructuration + indicatrice plus jolie

In this section, we first provide evidence which indicates that the main assumption made by papers which
found the previous best known lower bounds for weak Schur numbers using a computer may not be correct. 
This is done primarily by studying \(\WS(6)\). Then, in an effort to eliminate irrelevant search spaces, we 
obtain stronger results than those previously known for \(\WS (5)\) while gaining several orders of magnitude 
in computation time by giving additional information to the SAT solver without losing in generality. In this 
section, we assume that the subsets are ordered.

\subsection{Former assumption appears to be wrong}
Rowley's new lower bound for \(\WS (6)\) (642) \cite{RowleyWS} was a significant improvement upon
the former best known lower bound (582) \cite{EliahouBook}. This previous lower bound has been found 
several times using a computer (often with Monte-Carlo methods) and by recursively making the assumption 
that a good partition for \(\WS (n+1)\) starts with a good partition for \(\WS (n)\) which is true for small
values of \(n\). Therefore, one may wonder whether the limiting factor are the assumptions or the 
methods used to search for partitions. It appears that the search space induced by these assumptions does 
not contain the optimal partitions.

@@

\par
The above theorem shows that the previous lower bound for \(\WS (6)\) is optimal in the search space considered by the
papers which found it.
Therefore, finding a partition of \([\![1,n]\!]\) in 6 weakly sum-free subsets for some \(n \geqslant 590\) which does
not have a template-like structure
would be extremely interesting since it could give indications on a new search space for improving lower bounds with a
computer. More generally,
it questions the search space previously used for finding lower bounds for \(\WS (n)\) with a computer. In particular, to
our knowledge every paper
that found the lower bound \(\WS (5) \geqslant 196\) used this assumption. Therefore one may wonder
if this actually a good lower
bound. In the next subsection, we give properties that a partition of \([\![1,197]\!]\) in 5 weakly sum-free subsets has
to verify.

\subsection{Weak Schur number five}
As explained in the previous subsection, the search space used for showing that \(\WS (5) \geqslant 196\) may not contain
optimal solution. In this subsection,
we give necessary conditions for a hypothetical partition of \([\![1,197]\!]\) in 5 weakly sum-free subsets using the
same type of methods as in the
previous subsection.

@@

\subsection{Conclusion on the search space}
As explained in the first subsection, the recursive assumption that a good partition for\(\WS (n+1)\) starts with a good partition for 
\(\WS (n)\) appears to be wrong. Given that no extensive search for \(\WS(5)\) has been conducted without making this assumption and 
that the size of the considered partitions is reasonable (the current lower bound is 196 and \(S(5) = 160\) was verified), it seems worth 
investigating this special case further. In a spirit of orientating this search, we then give necessary conditions on an hypothetical partition 
of \([\![1,197]\!]\) in 5 weakly sum-free subsets. Finding partitions that exceeds the lower bounds found with a computer (even if they 
do not exceed those obtained with templates) which are not as regular as those obtained with templates would be extremely interesing 
since it could designate a new search space for finding lower bounds with a computer.

>>>>

\qquad In this section, we first provide evidence which indicate that the main assumption made by papers which
found the previous best known lower bounds for weak Schur numbers using a computer may not be correct. 
This is done primarily by studying \(\WS(6)\). Then, in an effort to eliminate irrelevant search spaces, we 
obtain stronger results than those previously known for \(\WS (5)\) while gaining several orders of magnitude 
in computation time by giving additional information to the SAT solver without losing in generality. In this 
section, we assume that the subsets are ordered.

\subsection{Former assumption appears to be wrong}

\qquad Rowley's new lower bound for \(\WS (6)\) (642) \cite{RowleyWS} was a significant improvement upon
the former best known lower bound (582) \cite{EliahouBook}. This previous lower bound has been found 
several times using a computer (often with Monte-Carlo methods) and by recursively making the assumption 
that a good partition for \(\WS (n+1)\) starts with a good partition for \(\WS (n)\) which is true for small
values of \(n\). Therefore, one may wonder whether the limiting factor are the assumptions or the 
methods used to search for partitions. It appears that the search space induced by these assumptions does 
not contain the optimal partitions.

@@

\par
The above theorem shows that the previous lower bound for \(\WS (6)\) is optimal in the search space considered by the
papers which found it.
Therefore, finding a partition of \([\![1,n]\!]\) in 6 weakly sum-free subsets for some \(n \geqslant 590\) which does
not have a template-like structure
would be extremely interesting since it could give indications on a new search space for improving lower bounds with a
computer.

\par More generally, this theorem questions the search space previously used for finding lower bounds for \(\WS (n)\) 
with a computer. In particular, to our knowledge every paper that found the lower bound \(\WS (5) \geqslant 196\) 
used this assumption. Therefore one may wonder if this is actually a good lower bound. In the next subsection, we give 
properties that a hypothetical partition of \([\![1,197]\!]\) in 5 weakly sum-free subsets has to verify.

\subsection{Weak Schur number five}

\qquad As explained in the previous subsection, the search space used for showing that \(\WS (5) \geqslant 196\) may not contain
optimal solution. In this subsection,
we give necessary conditions for a hypothetical partition of \([\![1,197]\!]\) in 5 weakly sum-free subsets using the
same type of methods as in the
previous subsection.

@@

\subsection{Conclusion on the search space}

\qquad As explained in the first subsection, the recursive assumption that a good partition for\(\WS (n+1)\) starts with a good partition for 
\(\WS (n)\) appears to be wrong. Given that no extensive search for \(\WS(5)\) has been conducted without making this assumption and 
that the size of the considered partitions is reasonable (the current lower bound is 196 and \(S(5) = 160\) was verified), it seems worth 
investigating this special case further. In a spirit of orientating this search, we then give necessary conditions on an hypothetical partition 
of \([\![1,197]\!]\) in 5 weakly sum-free subsets. Finding partitions that exceeds the lower bounds found with a computer (even if they 
do not exceed those obtained with templates) which are not as regular as those obtained with templates would be extremely interesing 
since it could designate a new search space for finding lower bounds with a computer.
